A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast

### 2021-10-24 analysis/paper/paper.Rmd
Moved to DTM 10. Started process of interpolating raster where there's railway/highway

# Introduction
The close association between Norwegian Stone Age sites located in coastal regions and prehistoric shorelines was firmly established by the geologist W.C. Brøgger and his son, the archaeologist A.W. Brøgger, in their respective theses of 1905 (see also @maccurdy1906). One of the major implications of this finding is that relative sea-level change can be used for the dating of the sites. If most Stone Age sites were situated on or close to the contemporary shoreline, a reconstruction of the trajectory of shoreline displacement can be used to date the original occupation of the sites based on their elevation relative to the present day sea level. The method of shoreline dating has consequently been fundamental to Stone Age research in Norway ever since, and is used both independently and to compliment the dating of sites where other sources of temporal data such as typological indicators or radiometric dates are either limited or not available. Given the coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating by means of shoreline displacement is often the only and most precise and accurate method by which one can hope to date the sites. Dating by means of shoreline displacement is consequently fundamental to our understanding of the Norwegian Stone Age. This is both because shoreline dating is fundamental to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shore-line. Consequently, adherence or deviation from this pattern has major implications for the socio-economic foundations of the societies in question.

Despite the fundamental role of shoreline-dating in Norway, there have only been conducted relatively small scale investigations into the applicability of dating by means of shore-line displacement, and the relationship has only been evaluated using what could be termed relatively coarse methods. The aim of this paper is to provide a systematic and comprehensive review of the degree to which radiocarbon dates correspond with the dates informed by our current knowledge of shoreline displacement in a larger area of south-eastern Norway, using a more refined methodological approach. The goal is to quantify the degree to which the assumption of shore-bound settlement holds through the Stone Age. This can be used both to inform the certainty with which one can depend on dating by means of shoreline displacement and its precision. It can also help identify societal shifts if there is chronological variation in the degree to which the sites are shore-bound, and any such variation can potentially also have implications for how archaeological investigations should be directed at different altitudes in the region. As presented in more detail below, this problem involves the combined evaluation of three major components. One is the questions of when the sites were in use, the second pertains to the reconstruction of the contemporaneous sea-level, and the third follows from the fact that the relation between site and shore-line is inherently a spatial question. A Monte Carlo simulation-based approach was adopted to integrate all three dimensions in the analysis. While this approach is only used to inform the question of the distance between sites and shorelines here, this method can be extended to a wide range of use-cases where one needs to evaluate and visualise the relationship between archaeological phenomena, the prehistoric shore-line, and the uncertainty inherent in this reconstruction.

@@

The most common way to evaluate whether or not a site would have been shorebound is typically through an evaluation of typological indicators in the assemblages and its correspondence with available shoreline displacement curves. If this corresponds then it is common to err on the side of a shorebound site location, even if the time-span of the dates provided by the typological indicators are too wide to decisively verify this. In the case that radiocarbon dates of site features are available, this is typically evaluated by means of a visual evaluation of of the overlap between shoreline displacement curves and the radiocarbon probability distributions. Another line of reasoning has been to combine these approaches with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would have been sensisble if the site was *not* shore bound [e.g. @jaksland2014]. This can for example pertain to accessibility. If the site is situated on a ledge in a steep and jagged area of the present day landscape, it would make intuitive sense that the location was used when the ocean reached the elevation of the ledge and the site would have been accessible by means of watercraft. This line of reasoning is often also used in survey projects which typically do not produce material suitable for radiocarbon dating. The evaluation of the degree to which sites were shorebound is therefore typically conducted on a site-by-site or project-by-project basis, meaning the fundamental basis for assuming shore bound site location is largely based on a mosaic of smaller scale investigations. Although it appears that the arguments for such site locations is sensible and for a large part appear to be correct, few attempts at a comprehensive evaluation and quantification of this tendency has been done.

>>>>

# Introduction
The relative sea-level fall that characterises most areas of Fennoscandia is fundamental to its archaeology, as archaeological phenomena that were originally shore-bound can be dated with reference to relative sea-level change . This follows from the fact that if the phenomena were situated on or close to the contemporary shoreline, a reconstruction of the trajectory of shoreline displacement can be used to date the original occupation of the sites based on their elevation relative to the present day sea level.

Post-glacial isostatic rebound leading to shoreline fall

The close association between Norwegian Stone Age sites located in coastal regions and prehistoric shorelines was firmly established by the geologist W.C. Brøgger and his son, the archaeologist A.W. Brøgger, in their respective theses of 1905 (see also @maccurdy1906). One of the major implications of this finding is thus that relative sea-level change can be used for the dating of the sites. The method of shoreline dating has consequently been fundamental to Stone Age research in Norway ever since. The method is used both independently and to compliment the dating of sites where other sources of temporal data such as typological indicators or radiometric dates are either limited or not available. Given the coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating by means of shoreline displacement is often the only and most precise and accurate method by which one can hope to date the sites. Dating by means of shoreline displacement is consequently fundamental to our understanding of the Norwegian Stone Age. However, this is both because shoreline dating is fundamental to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shore-line. Consequently, adherence or deviation from this pattern has major implications for the socio-economic foundations of the societies in question.

Despite the fundamental role of shoreline-dating in Norway, there have only been conducted relatively small scale investigations into the applicability of dating by means of shore-line displacement, and the relationship has only been evaluated using what could be termed relatively coarse methods. The aim of this paper is to provide a systematic and comprehensive review of the degree to which radiocarbon dates correspond with the dates informed by our current knowledge of shoreline displacement in a larger area of south-eastern Norway, using a more refined methodological approach. The goal is to quantify the degree to which the assumption of shore-bound settlement holds through the Stone Age. This can be used both to inform the certainty with which one can depend on dating by means of shoreline displacement as well as its precision. The investigation can also help identify societal shifts, if there is chronological variation in the degree to which the sites are shore-bound, and any such variation can potentially also have implications for how archaeological investigations should be directed at different altitudes in the area of analysis. As presented in more detail below, this problem involves the combined evaluation of three major components. One is the questions of when the sites were in use, the second pertains to the reconstruction of the contemporaneous sea-level, and the third follows from the fact that the relation between site and shore-line is inherently a spatial question. A Monte Carlo simulation-based approach was adopted to integrate all three dimensions in the analysis. The approach is only used to inform the question of the distance between sites and shorelines here, but this method can also be extended to a wide range of use-cases where one needs to evaluate and visualise the relationship between archaeological phenomena, the prehistoric shore-line, and the uncertainty inherent in this reconstruction.

@@

The most common way to evaluate whether or not a site would have been shorebound is typically through an evaluation of typological indicators in the assemblages and its correspondence with available shoreline displacement curves. If this corresponds then it is common to err on the side of a shorebound site location, even if the time-span of the dates provided by the typological indicators are too wide to decisively verify this. In the case that radiocarbon dates of site features are available, this is typically evaluated by means of a visual evaluation of the overlap between shoreline displacement curves and the radiocarbon probability distributions. Another line of reasoning has been to combine these approaches with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would have been sensible if the site was *not* shore bound [e.g. @jaksland2014]. This can for example pertain to accessibility. If the site is situated on a ledge in a steep and jagged area of the present day landscape, it would make intuitive sense that the location was used when the ocean reached the elevation of the ledge and the site would have been accessible by means of watercraft. This line of reasoning is often also used in survey projects which typically do not produce material suitable for radiocarbon dating. The evaluation of the degree to which sites were shorebound is therefore typically conducted on a site-by-site or project-by-project basis, meaning the fundamental basis for assuming shore bound site location is largely based on a mosaic of smaller scale investigations. Although it appears that the arguments for such site locations is sensible and for a large part appear to be correct, few attempts at a comprehensive evaluation and quantification of this tendency has been done.
